---
title: "Critical Loads Summary for `r params$unit`"
# title: "\vspace{-1.5cm}{Critical Loads Summary \break for `r params$unit`}"
header-includes:
- \usepackage{hyperref} \hypersetup{pdfstartview={XYZ null null 1.00}, colorlinks
  = true, linkcolor = red, urlcolor = red}
- \usepackage{float} \floatplacement{figure}{H}
- \usepackage[font=small,skip=0pt]{caption}
- \usepackage{wrapfig}
- \usepackage{booktabs}
- \usepackage{longtable}
params:
  unit: 'Bridger-Teton National Forest'
output:
  pdf_document:
    latex_engine: lualatex
  # html_document:
  #   df_print: paged
toccolor: black
---

```{r setup-options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '')
```

```{r load-csv, include=FALSE}
# x <- read.csv('path/to/file.csv', strip.white=TRUE)
```

```{r setup-functions, include=FALSE}
### set graphical parameters
`set_par` <- function (panels = 1, pty = 's', ...) {
    mgp <- c(2, 0.4, 0)
    mar <- c(4, 4, 0.5, 0.5)
    oma <- c(0, 0, 0, 0)
    auto_rowcol <- function(n = panels) {
        if (n <= 3) 
            c(1, n)
        else if (n <= 6) 
            c(2, (n + 1)%/%2)
        else if (n <= 12) 
            c(3, (n + 2)%/%3)
        else c(ceiling(n/(nr <- ceiling(sqrt(n)))), nr)
    }
    mfrow <- auto_rowcol()
    if (panels > 4) 
        panels <- 4
    switch(as.character(panels), `4` = par(mfrow = mfrow, 
        mgp = mgp, mar = mar, pty = pty, oma = oma, bty = 'L', 
        las = 1, cex.lab = 1.2, tcl = -0.2, ...), `3` = par(mfrow = mfrow, 
        mgp = mgp, mar = mar, pty = pty, oma = oma, bty = 'L', 
        las = 1, cex.lab = 1.4, cex.axis = 1.2, tcl = -0.2, ...), 
        `2` = par(mfrow = mfrow, mgp = mgp, mar = mar, 
            pty = pty, oma = oma, bty = 'L', las = 1, cex.axis = 0.85, 
            tcl = -0.2, ...), `1` = par(mfrow = mfrow, 
            mgp = mgp, mar = mar, pty = pty, oma = oma, bty = 'L', 
            las = 1, cex.axis = 0.85, tcl = -0.2, ...))
}
# ### get some value
# `get_n` <- function (pick, ...) {
#         n <- x[x$pid==pick, 'n_plots']
#         if (any(!is.finite(n))) c(NA) else n
# }
# ### table function with 5-y mean
# `get_tab` <- function (pick, xvar, col.names, caption, format){
#         tab <- x[which(x$pid==pick), xvar]
#         if (missing(col.names)) col.names <- xvar
#         if (missing(caption))   caption   <- NULL
#         if (missing(format))    format    <- 'pandoc'
#         j   <- xvar[xvar != 'tenyr'] # dont consider year column
#         tab <- tab[rowSums(is.na(tab[,j])) != ncol(tab[,j]),]
#         tab[tab %in% c('NA','***NA***','NaN','***NaN***')] <- NA 
#         knitr::kable(x         = tab,
#                      format    = format,
#                      row.names = F,
#                      align = c('l', rep('r',NCOL(tab)-1)),
#                      longtable = T,
#                      booktabs  = T,
#                      caption   = caption,
#                      col.names = col.names)
# }
```

<!-- \vspace*{\fill} -->
<!-- \begin{wrapfigure}{l}{1.8cm} -->
<!-- \includegraphics{usfs.png} -->
<!-- \end{wrapfigure} -->

<!-- \   -->

<!-- Air Resource Management Program   -->
<!-- USDA Forest Service   -->
<!-- 1400 Independence Ave SW   -->
<!-- Washington, DC 20250   -->

<!-- \   -->

<!-- \newpage -->

<!-- ####    ####    ####    ####    ####    ####    ####    ####    #### -->
<!-- ####    ####    ####    ####    ####    ####    ####    ####    #### -->
<!-- ####    ####    ####    ####    ####    ####    ####    ####    #### -->



## Motivation

A critical load (CL) is the atmospheric deposition level expected to result in harmful ecosystem changes.  Exceedance is the degree to which locally observed deposition goes beyond an established CL.  See **[critical loads we used](#critical-loads-we-used)**).

### Estimated deposition range for `r params$unit`

Minimum total N deposition: `xxxx` kg ha^-1^ y^-1^  
Maximum total N deposition: `xxxx` kg ha^-1^ y^-1^  

### Critical loads

Alpine minimum: \tab z  
Alpine maximum: \tab z  
Herb richness minimum: \tab z  
Herb richness maximum: \tab z  
Herb richness 10^th^ percentile: \tab z  
Herb richness median: \tab z  
Tree growth minimum: \tab z  
Tree growth median: \tab z  
Tree survival minimum: \tab z  
Tree survival median: \tab z  
Lichen minimum: \tab z  
Lichen maximum: \tab z  
Mycorrhizal fungi minimum: \tab z  
Mycorrhizal fungi maximum: \tab z  
Nitrate Leaching minimum: \tab z  
Nitrate Leaching maximum: \tab z  
Aquatic minimum: \tab z  
Aquatic maximum: \tab z  





```{r critical_loads_tab}
# tab <- data.frame(Metric = c('Species richness',
#                              'Sensitive species richness',
#                              'Forage lichen abundance',
#                              'Cyanolichen abundance',
#                              'Comm composition shifts'),
#                   Ndep = c(3.5, 3.1, 1.9, 1.3, 1.5),
#                   Sdep = c(6.0, 2.5, 2.6, 2.3, 2.7))
# knitr::kable(tab, format='pandoc', row.names = F, longtable = T,
#              booktabs  = T,
#              caption   = 'Critical loads used in this report (kg ha^-1^ y^-1^).',
#              escape    = F,
#              col.names = c('Lichen metric', 'N deposition', 'S deposition'))
```


<!-- \begin{wrapfigure}{r}{5.0cm} -->
<!-- \includegraphics{bothlichens.png} -->
<!-- \end{wrapfigure} -->



## References

>Bowman, W. D., J. Murgel, T. Blett, and E. Porter. 2012. Nitrogen critical loads for alpine vegetation and soils in Rocky Mountain National Park. Journal of Environmental Management 103:165-171.

>Horn, K. J., R. Q. Thomas, C. M. Clark, L. H. Pardo, M. E. Fenn, G. B. Lawrence, S. S. Perakis, E. A. H. Smithwick, D. Baldwin, S. Braun, A. Nordin, C. H. Perry, J. N. Phelan, P. G. Schaberg, S. B. St. Clair, R. Warby, and S. Watmough. 2018. Growth and survival relationships of 71 tree species with nitrogen and sulfur deposition across the conterminous U.S. PLoS ONE 13:e0205296.

>Simkin, S. M., E. B. Allen, W. D. Bowman, C. M. Clark, J. Belnap, M. L. Brooks, B. S. Cade, S. L. Collins, L. H. Geiser, F. S. Gilliam, S. E. Jovan, L. H. Pardo, B. K. Schulz, C. J. Stevens, K. N. Suding, H. L. Throop, and D. M. Waller. 2016. Conditional vulnerability of plant diversity to atmospheric nitrogen deposition across the United States. Proceedings of the National Academy of Sciences 113:4086-4091.



<!-- END -->
